Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-05-28 paper.tex
more edits
more response edits

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as mulitple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

The proof of Theorem~\ref{thm:consistency} relies on Lemmas \ref{lem:convtoexpGlamL} to~\ref{lem:GLtoG0}, which can be found in the appendix.
Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:LSallparams} we obtain $LS$ minimised with respect to the data symbols,
 \begin{align*}
LS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$,
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)).
\end{equation}
%where $\Re(\cdot)$ denotes the real part.  %Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  
Substituting this expression into $LS(\rho, \theta)$ gives $LS$ minimised with respect to $\rho$ and the data symbols,
\[
LS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
%We again abuse notation by reusing $LS$, but this should not cause confusion as $LS(\rho,\theta)$ and $LS(\theta)$ are easily told apart by their inputs.  
By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  %In this case we would replace $\theta$ with $\fracpart{\theta + \phi}$ and $\hat{\rho}(\theta)$ with $-\hat{\rho}(\theta)$ and observe that
%\[
%LS(\rho,\theta) = LS(-\rho, \fracpart{\theta + \pi})
%\]
The least square estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $LS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ with no constraints required.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.  As expected, the figures indicate the estimator to become more accurate as the number of pilot symbols is increased.  An interesting property distriguishing the case when no pilot symbols exist is the slope of the MSE curve at low SNR.  When a finite proportion of pilot symbols exist the slope at low SNR is similar to that when all symbols are pilots.  The slope is steeper in the case that no pilot symbols exist.

@@

Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} show the mean square error of the amplitude estimator when $M=2,4,8$ and $L=4096$ and when the number of pilot symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the squared bias $\rho_0^2\big(1 - G(0)\big)^2$ plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\ref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When the SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases so the bias is significant when SNR is small.  Figure~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.

>>>>

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as multiple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

The proof of Theorem~\ref{thm:consistency} relies on Lemmas \ref{lem:convtoexpGlamL} to~\ref{lem:GLtoG0}, which can be found in the appendix.
Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:LSallparams} we obtain $LS$ minimised with respect to the data symbols,
 \begin{align*}
LS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$,
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)).
\end{equation}
%where $\Re(\cdot)$ denotes the real part.  %Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  
Substituting this expression into $LS(\rho, \theta)$ gives $LS$ minimised with respect to $\rho$ and the data symbols,
\[
LS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
%We again abuse notation by reusing $LS$, but this should not cause confusion as $LS(\rho,\theta)$ and $LS(\theta)$ are easily told apart by their inputs.  
By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  %In this case we would replace $\theta$ with $\fracpart{\theta + \phi}$ and $\hat{\rho}(\theta)$ with $-\hat{\rho}(\theta)$ and observe that
%\[
%LS(\rho,\theta) = LS(-\rho, \fracpart{\theta + \pi})
%\]
The least squares estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $LS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ with no constraints required.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar conclusions are made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.  As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, at low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.

@@

Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} show the mean square error of the amplitude estimator when $M=2,4,8$ and $L=4096$ and when the number of pilot symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the squared bias $\rho_0^2\big(1 - G(0)\big)^2$ plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\ref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\] 
When the SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases so the bias is significant when SNR is small.  Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.

### 2014-05-28 response2.tex
more response edits

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data-aided case \dots'' $\to$ bounds for the common scenario where there is a mixture of pilot and data symbols can be found in ``N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005.'' \\
\textbf{Response:} We have added this reference and have removed the sentences begining ``To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case \dots''. Note that the model considered in the paper by Noels~et.~al.~\cite{Noels_crbs_with_pilots_2005} treats the amplitude as fixed and known and includes a frequency offset term.  Thus, the bounds are not immediately applicable in our setting and we are not able to meaningfully compare these bounds with our simulation results.  %For this reason with have kept with the bound described in~\cite{950345} as in the previous version of the manuscript.

@@

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:} Reviewer 1 comment~\eqref{eq:rev1:multantenna} wanted ``antenna'' and we have used this spelling.

@@

\item \textbf{Comment:} p. 5 (and further throughout the text): ``pilots symbols" (with `s') and ``pilot symbols'' (without `s') are both used $\to$ use only 1, I prefer ``pilot symbols'' \\
\textbf{Response:} All occurences of ``pilots symbols'' have been changed to ``pilot symbols''.

@@

\item \textbf{Comment:} before (6): drop the 'to' \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} I would prefer to see comma's between two subindices; so, `$k,i$' rather than`'$ki$' \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} p. 14: ``Equivalently $\hat\theta$ is the maximiser of \dots'' $\to$ could you please provide some further explanation (is the maximum of $\Re(Z(\theta))$ never negative? and why) \\
\textbf{Response:} The least squares phase estimator is the minimiser of
\[
LS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
Because the least squares amplitude estimator $\hat{\rho} = \Re(Z(\hat{\theta}))$ is positive by definition, we are only interested in those values of $\theta$ that correspond with a positive amplitude estimator, that is, such that
\[
\hat{\rho}(\theta) = \Re(Z(\theta)) > 0.
\]
Thus, the least squares phase estimator $\hat{\theta}$ must be such that
\[
\hat{\rho} = \Re(Z(\hat{\theta})) > 0.
\]
This is explained in the text by the sentences:

``By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  The least square estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $LS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ with no constraints required.''

@@

\item \textbf{Comment:} in general, I find the numerical results section rather short \\
\textbf{Response:} We have greatly expanded the simulation section as detailed in the responses that follow.

@@

``As expected, the figures indicate the estimator to become more accurate as the number of pilot symbols is increased.  An interesting property distriguishing the case when no pilot symbols exist is the slope of the MSE curve at low SNR.  When a finite proportion of pilot symbols exist the slope at low SNR is similar to that when all symbols are pilots.  The slope is steeper in the case that no pilot symbols exist.''

@@

\item \textbf{Comment:} p. 18, first sentence: a qualitative explanation for the fact the theory and simulations diverge at low SNR would be appreciated  \\
\textbf{Response:} BLERG: Added a reference to Tazras crb paper.

@@

\textbf{Response:}

@@

@@

\item \textbf{Comment:}  There are many formulas (starting from page 11) with the notation $\fracpart{x}$ meaning ``$x$ taken modulo $2\pi/M$''. These angle brackets don't strike the eye and can be easily confused with ordinary parentheses. To make them much more visible, the authors could use the same approach as for ``$x$ taken modulo $2\pi$'' - by using corresponding subscript, i.e. use $\fracpart{x}_{2\pi/M}$ instead of $\fracpart{x}$. \\
\textbf{Response:} We request not to make this change.  There are two reasons.  The first is that this notation leads to situations where there are subscripts withing superscripts.  For example, in the proof of Theorem 1 we would have expressions of the form
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}_{2\pi/M}}
\]
and in the appendix we would obtain expressions such as
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}_{2\pi/M}}\right)
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}_{2\pi/M}} - R_1 e^{j\fracpart{\Phi_1}_{2\pi/M}}  \big].
\]
We feel that these are more difficult to parse that the existing simpler
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right),
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]

@@

\input{paper.bbl}

>>>>

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data-aided case \dots'' $\to$ bounds for the common scenario where there is a mixture of pilot and data symbols can be found in ``N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005.'' \\
\textbf{Response:} We have added this reference and have removed the sentences beginning ``To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case \dots''. Note that the model considered in the paper by Noels~et.~al.~\cite{Noels_crbs_with_pilots_2005} treats the amplitude as fixed and known and includes a frequency offset term.  Thus, the bounds are not immediately applicable in our setting and we are not able to meaningfully compare these bounds with our simulation results.  %For this reason with have kept with the bound described in~\cite{950345} as in the previous version of the manuscript.

@@

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:} Reviewer 1 comment~\eqref{eq:rev1:multantenna} wanted ``antenna'' and we have used this spelling. ``Multiple'' is now spelt correctly.

@@

\item \textbf{Comment:} p. 5 (and further throughout the text): ``pilots symbols" (with `s') and ``pilot symbols'' (without `s') are both used $\to$ use only 1, I prefer ``pilot symbols'' \\
\textbf{Response:} All occurrences of ``pilots symbols'' have been changed to ``pilot symbols''.

@@

\item \textbf{Comment:} before (6): drop the `to' \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} I would prefer to see comma's between two subindices; so, `$k,i$' rather than`$ki$' \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} p. 14: ``Equivalently $\hat\theta$ is the maximiser of \dots'' $\to$ could you please provide some further explanation (is the maximum of $\Re(Z(\theta))$ never negative? and why) \\
\textbf{Response:} The least squares phase estimator is the minimiser of
\[
LS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
Because the least squares amplitude estimator $\hat{\rho} = \Re(Z(\hat{\theta}))$ is positive by definition, we are only interested in those values of phase $\theta$ that correspond with a positive amplitude estimator, that is, such that
\[
\hat{\rho}(\theta) = \Re(Z(\theta)) > 0.
\]
Thus, the least squares phase estimator $\hat{\theta}$ must be such that
\[
\hat{\rho} = \Re(Z(\hat{\theta})) > 0.
\]
This is explained in the text by the sentences:

``By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  The least squares estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $LS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ with no constraints required.''

@@

\item \textbf{Comment:} in general, I find the numerical results section rather short \\
\textbf{Response:} We have expanded the simulation section as detailed in the responses that follow.

@@

``As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, at low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.''

@@

\item \textbf{Comment:} p. 18, first sentence: a qualitative explanation for the fact the theory and simulations diverge at low SNR would be appreciated  \\
\textbf{Response:} The following text has been included in Section~\ref{sec:simulations}:

``As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar conclusions are made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.''

@@

\textbf{Response:} Figure~\ref{fig:biasvarcomp} in this document displays such a plot.  The number of symbols is set to $L=4096$.  The solid line shows that asymptotic squared bias 
\[
\rho_0^2\big(1 - G(0)\big)^2
\]
and the dashed line in the asymptotic variance
\[
\rho_0^2\frac{p B_1 + d B_2}{L}.
\]
The number of pilots consider is $\abs{P}=0$ and $\abs{P}=L/2$.  The obvious relationship is observed: the bias dominates the variance at low SNR and the variance dominates the bias at high SNR.  We believe that this plot is not particularly informative and have not included it in the manuscript.

@@

\begin{figure}[tp] 
	\centering  
		\includegraphics[]{code/data/biasvarplot-1.mps}
		\caption{Plot of asymptotic bias squared (dashed) and variance (solid) of the amplitude estimator for QPSK with $L=4096$.}
		\label{fig:biasvarcomp}
\end{figure}

@@

\item \textbf{Comment:}  There are many formulas (starting from page 11) with the notation $\fracpart{x}$ meaning ``$x$ taken modulo $2\pi/M$''. These angle brackets don't strike the eye and can be easily confused with ordinary parentheses. To make them much more visible, the authors could use the same approach as for ``$x$ taken modulo $2\pi$'' - by using corresponding subscript, i.e. use $\fracpart{x}_{2\pi/M}$ instead of $\fracpart{x}$. \\
\textbf{Response:} We respectfully request not to make this change.  There are two reasons.  The first is aesthetic.  For example, in the proof of Theorem 1 we would have expressions of the form
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}_{2\pi/M}}
\]
and in the appendix we would obtain expressions such as
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}_{2\pi/M}}\right)
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}_{2\pi/M}} - R_1 e^{j\fracpart{\Phi_1}_{2\pi/M}}  \big].
\]
We feel that these are more difficult to parse that the existing simpler
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right),
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]

@@

{
\small
\input{paper.bbl}
}
